---
title: "hw0"
author: "Nancy Zha"
format: html
editor: visual
execute:
  echo: false
---

```{r}
library(knitr)
```

(1) Write a Quarto document that defines variables $a=1, b=-1, c=-2$ and print out the solutions to $f(x) = ax^2+bx+c=0$. Do not report complex solutions, only real numbers.

```{r}
quadratic_fun <- function(a, b, c){
  # a*x^2 + b*x + c = 0
  #print(paste0('Trying to solve solutions for:', a, 'x^2 +', b, 'x +', c))
  disc = b^2 - 4*a*c
  if (disc > 0){
    x1 = (-b + sqrt(disc))/(2*a)
    x2 = (-b - sqrt(disc))/(2*a)
    return (paste0('The solutions are: ', x1,' and ', x2))
  }
  else if (disc == 0){
    x = -b / (2*a)
    return (paste0('The solution is: ', x))
  }
  else {
    return('No real number solution.')
  }
}
```

```{r}
#quadratic_fun(a = 1, b = -1, c = -2)
```

(2) Include a graph of $f(x)$ versus $x$ for $x \in (-5,5)$.

This is how you make a plot of a quadratic function:

(3) Generate a PDF report using knitr. Do not show the R code, only the solutions and explanations of what the reader is seeing.

(4) Erase the PDF report and reproduce it but this time using $a=1, b=2, c=5$.

```{r}
#quadratic_fun(a = 1, b = 2, c = 5)
```

(5) Erase the PDF report and reproduce it but this time using $a=1, b=3, c=2$.

Answer: I solved the quadratic equation by creating a function using quadratic formula $x = \frac{-b \pm\sqrt{b^2 - 4ac}}{2a}$ for the discriminant $b^2 - 4ac$. 

Three conditions are considered: 1. discriminant\>0, 2 real roots. 2. discriminant\<0, no real roots. 3. discriminant=0, 2 equal roots. In our case, the discriminant is larger than 0 so we have two different solutions. Plot shows parabola of $f(x) = x^2+3x+2$ for $x \in (-5,5)$ and the solutions are at intersection of parabola and the $y=0$ line, which are $x_1=-1$ and $x_2=-2$.

```{r}
quadratic_fun(a = 1, b = 3, c = 2)
```

```{r}
a <- 1 
b <- 3
c <- 2
x <- seq(-5, 5, length = 200)
plot(x, a*x^2 + b*x + c, type = "l", ylab = "f(x)")
abline(v = -1)
text(x = -1, y = 2, "x1")
abline(v = -2)
text(x = -2, y = 2, "x2")
abline(h = 0, lty = 2)
```

(6) Create an HTML page with the results for this last set of values, but this time showing the code.
